### 2013-10-15 Publications/2013-TMI/04-methods.tex
Making mathematical notation homogeneous
Still WIP

\subsection{Proposed model}
\label{sec:methods_map}
In this paper we formulate the joint registration-segmentation
problem as follows. We look for a spatial mapping (with, typically,
$n=3$ for the 3-dimensional case):
\begin{align}
U\colon T \subset \mathbb{R}^n &\to D \subset \mathbb{R}^n \notag\\
\mathbf{x} &\mapsto \mathbf{x}' =\mathbf{x}+u(\mathbf{x}),
\label{eq:deformation_field}
\end{align}
such that the known contours in anatomical space $T$ optimally segment
the diffusion space $D$.

In a Bayesian framework, different such mappings $U$ between ``T1-''
and ``\gls*{dwi}-'' spaces are evaluated based on their posterior
probability given the observed data $F$. Using the Bayes' rule, this
posterior \emph{likelihood} can be computed as:
\begin{equation}
P(U \mid F,\Gamma_{ij}) = \frac{P(F \mid U,\Gamma_{ij})\, P(U)}{P(F)},
\label{eq:bayes_rule}
\end{equation}
where $P(F \mid U,\Gamma_{ij})$ is the data-likelihood, and
$\Gamma_{ij}$ are a set of contours in T1-space that correspond to
two interfacing regions $\Omega$, such that 
$\Gamma_{ij} = \partial \Omega_i \cap \partial \Omega_j$.

We are interested in a mapping $\hat{U}$ which maximizes this
posterior probability (\gls*{map} estimate, \citep{bishop_pattern_2006}):
\begin{align}
\hat{U} &= \underset{U}{\argmax} \left\{ P(U \mid F,\Gamma_{ij}) \right\} \notag\\
 &= \underset{U}{\argmax} \left\{ P(F \mid U,\Gamma_{ij})\, P(U) \right\}.
\label{eq:map_u}
\end{align}

@@

\begin{equation}
P(F \mid U,\Gamma_{ij}) = 
\underset{\mathbf{x}\in \Omega}{\prod} P\left( F(\mathbf{x}') \mid U(\Gamma_{ij}) \right).
\label{eq:bayes_aposteriori}
\end{equation}

Further, within each region $\Omega_l$ defined by the contours $\Gamma$
trough duality, we assume the features to be i.i.d.:
\begin{equation}
P\left( F(\mathbf{x}') \mid U,\Gamma_{ij} \right) = p_l( F(\mathbf{x}')) \quad
\forall \mathbf{x} \in \Omega_l.
\label{eq:likelihood}
\end{equation}

@@

\begin{equation}
p_l( F(\mathbf{x}') ) = \mathcal{N} ( f \mid \Theta_l ).
\label{eq:multivariate_normal}
\end{equation}

\todo{I modified this slightly}
Thus, defining the feature observed as $\mathbf{f}'=F(\mathbf{x}')$ and 
the squared \emph{Mahalanobis distance} of $\mathbf{f}'$ with respect
of the descriptors of region $l$ as
$\Delta^2_l (\mathbf{f}' \mid \Theta_l ) = (\mathbf{f}' - \boldsymbol{\mu}_l)^T \, \boldsymbol{\Sigma}^{-1}_l \, (\mathbf{f}' - \boldsymbol{\mu}_l)$,
we write:
\begin{align}
P( F \mid U, \Omega ) &= \underset{l}{\prod} \underset{\mathbf{x} \in \Omega_l}{\prod}
\mathcal{N} ( f \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) \notag\\
&= \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}  \Delta^2_l (\mathbf{f}' \mid \Theta_l ) \right)}}.
\label{eq:pdf}
\end{align}

We choose a Thikonov regularization prior as follows:

\begin{equation*}
P(U) = \underset{\mathbf{x}}{\prod} p(u)
\end{equation*}
with further:
\begin{align*}
p(u) &= p_0(u) \, p_1(u) \notag\\
p_0(u) &= \mathcal{N}( u \mid 0, A^{-1}) \notag\\
p_1(u) &= \mathcal{N}( Du \mid 0, B^{-1}),
\end{align*}
that consider that the distortion and its gradient have zero
mean and variance governed by $A$ and $B$. Since the anisotropy
is technically aligned with the imaging axes, these can
be simplified:
\begin{align}
p_0(u) &= \underset{\mathbf{x}}{\sum} \mathcal{N}( u(\mathbf{x}) \mid 0, \frac{1}{\sqrt{a_{x,y,z}}}) \notag\\
p_1(u) &= \underset{\mathbf{x}}{\sum} \mathcal{N}( \nabla_{x,y,z}u(\mathbf{x}) \mid 0, \frac{1}{\sqrt{b_{x,y,z}}})
\label{eq:priors}
\end{align}

Finally, we can turn the \gls{map} problem into 
a variational one
applying the following log-transform:
\begin{align}
E(F \mid U) &= -\log \underset{l}{\prod}
\underset{\mathbf{x} \in \Omega_l}{\prod} 
p_l( \mathbf{f}')\,p_0( u(\mathbf{x}))\,p_1( u(\mathbf{x})) \notag\\
&= C + \underset{l}{\sum}
\underset{\mathbf{x} \in \Omega_l}{\sum}
\Delta^2_l (\mathbf{f}' \mid \Theta_l ) \notag\\
&+ \underset{x,y,z}{\sum} a_d u^2_d
+ \underset{x,y,z}{\sum} b_d \|\nabla u_d\|^2,
\label{eq:energy}
\end{align}
that is the dual expression to the energy functional corresponding
to a discrete \gls*{acwe} framework \citep{chan_active_2001}
with anisotropic regularization as studied in
\citep{nagel_investigation_1986}.

{\color{red} {Using the region descriptors derived in \autoref{sec:methods_map}, we propose
an \gls{adf}-like, \gls{acwe}-based, piece-wise constant, image segmentation
model (where the unknown is the deformation field)
\cite{chan_active_2001} with the energy functional obtained in 
\eqref{eq:final_map_energy}. This inverse problem is ill-posed
\cite{bertero_ill-posed_1988,hadamard_sur_1902}.
In order to account for deformation field regularity and to render the 
problem well-posed, we include limiting and regularization terms into 
the energy functional \cite{morozov_linear_1975,tichonov_solution_1963}:
%
\begin{multline}
E(u) = \sum\limits_l \int_{\Omega'_l} \Delta^2_l (\mathbf{f}) \,d\mathbf{x} \\
+ \int u^T \, A \, u \, d\mathbf{x} + \int \tr\{(\nabla u^T)^T B (\nabla u^T)\} d\mathbf{x}
\label{eq:complete_energy}
\end{multline}
%
These regularity terms ensure that the segmenting contours in 
\gls{dwi} space are still close to their native shape. The model
easily allows to incorporate inhomogeneous and anisotropic 
regularization \cite{nagel_investigation_1986} to better regularize
the \gls{epi} distortion.}}

@@

\subsection{Numerical Implementation}
\label{sec:numerical_implementation}
%
Let us denote $\{c_i\}_{i=1 \ldots N_c}$ the nodes of one or several shape-prior
surface(s). In our application, precise tissue interfaces of interest 
extracted from a high-resolution, anatomically correct reference volume. 
On the other hand, we have a number of \gls{dwi}-derived features at each
voxel of the volume. Let us denote by $\mathbf{x}$ the voxel and $F(\mathbf{x}) = 
[ f_1, f_2, \ldots, f_N]^T(\mathbf{x})$ its associated feature vector.

\subsubsection{Deformation field}
\label{sec:deformation_field}
The transformation from ``T1-'' into ``\gls{dwi}-'' coordinate space is 
achieved through a dense deformation field $u(\mathbf{x})$, such that:
\begin{equation}
c_i' = U\{c_i\} = c_i + u(c_i),
\label{eq:nodes_tfm}
\end{equation}
where $U$ is defined in \eqref{eq:deformation_field}. Since the nodes of the anatomical 
surfaces might lay off-grid, it is required to derive $u(\mathbf{x})$ from a discrete 
set of parameters $\{u_k\}_{k=1 \ldots K}$. Densification is achieved through 
a set of associated basis functions $\psi_k$:
%
\begin{equation}
u(\mathbf{x}) = \sum_k \psi_k(\mathbf{x}) u_k
\label{eq:intp_kernel}
\end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
of degree 3 ($B_3$).
Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
$\psi$ by the actual kernel function, the transformation writes
%
\begin{equation}
c_i' = c_i + \sum_k \left[ u_k \, \underset{x,y,x}{\prod} B_3(c_{i,d} - \mathbf{x}_{k,d}) \right].
\label{eq:transformation}
\end{equation}

Based on the current estimate of the distortion $u$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$\mathbf{f}'$ of the regions defined by the priors in \gls{dwi} space.

\subsubsection{Gradient descent}
To find the minimum of the energy functional \eqref{eq:energy},
we propose a gradient-descent approach with respect to the underlying 
deformation field through the following partial differential equation:
\begin{equation}
\frac{\partial u(\mathbf{x},t)}{\partial t} \propto - \frac{\partial E(u)}{\partial u},
\label{eq:general_gradient_descent}
\end{equation}
with $t$ being an artificial time parameter of the contour
evolution, included to express the iterative character of the approach.
Now, we can introduce \eqref{eq:energy} in
\eqref{eq:general_gradient_descent}:
\begin{align}
\frac{\partial E(u)}{\partial u^d_k} &=
\frac{ \partial }{\partial u^d_k} \lbrace
C + \underset{l}{\sum}
\underset{\mathbf{x} \in \Omega_l}{\sum}
\Delta^2_l (\mathbf{f}' \mid \Theta_l ) \notag\\
&+ \underset{a}{\sum} a^d ({u^d})^2
+ \underset{a}{\sum} b^d \|\nabla u^d\|^2
\rbrace
\label{eq:gradient_descent}
\end{align}

Whereas related \glspl*{adf} introduced in \autoref{sec:methods_background}
make use of explicit level-set formulations to solve \eqref{eq:gradient_descent},
we alternatively use \emph{shape-gradients}
\cite{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}.
Let $r(\mathbf{x})$ be an ``arbitrary'' function over the image domain,
and $\Omega$ a certain image region with $\Gamma_{ij}$ its corresponding
outer boundary as defined in \autoref{sec:methods_map}.
We now derive the domain integral w.r.t. $t$:
\begin{equation*}
\frac{\partial}{\partial t} \int_\Omega r(\mathbf{x}') d\mathbf{x} = \int_\Omega \frac{\partial}{\partial t}r(\mathbf{x}') d\mathbf{x} - \int_{\Gamma_{ij}} r(\mathbf{x}') \left\langle \frac{\partial \Gamma_{ij} }{\partial t}, N_{\Gamma_{ij}}\right\rangle d\mathbf{x}
\end{equation*}
where $\left\langle\frac{\partial\Gamma_{ij}}{\partial t}, N_{\Gamma_{ij}}\right\rangle$ is 
the projection of the boundary movement on the unit inward normal $N_{\Gamma_{ij}}$. Assuming
that the region descriptors $\Theta_l$ vary slowly enough, we can consider
that $\frac{\partial}{\partial t} r(\mathbf{x}') = 0$ and thus:
\begin{equation}
\frac{\partial}{\partial t} \int_\Omega r(\mathbf{x}') d\mathbf{x} = 
- \int_{\Gamma_{ij}} r(\mathbf{x}') \left\langle \frac{\partial \Gamma_{ij} }{\partial t}, N_{\Gamma_{ij}}\right\rangle d\mathbf{x}
\label{eq:shape_gradients}
\end{equation}

Therefore, we can apply a discretized interpretation of \eqref{eq:shape_gradients}
to compute the data term in \eqref{eq:gradient_descent} as follows:
\begin{align}
\frac{\partial E_{data}(u)}{\partial u^d_k} &=
\underset{l}{\sum} \frac{ \partial }{\partial u^d_k} \lbrace
 \underset{\mathbf{x} \in \Omega_l}{\sum}
\Delta^2_l (\mathbf{f}') \rbrace \notag\\
&= \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right]
\left\langle \frac{\partial \mathbf{c}'}{\partial {u^a_k}'}, \mathbf{n_i}'\right\rangle,
\label{eq:gradient_wshape}
\end{align}
in this case, the formulation has been adapted to the non-binary case, $ij$
being any pair of neighboring regions, and $\Gamma_{ij}$ the contour separating
them such that $\mathbf{x}' = \mathbf{c}' \in\Gamma_{ij} \iff \mathbf{x}\in \partial\Omega_i \cap \partial\Omega_j$
and $\mathbf{n_i}'$ is the unit inward normal to the contour at $c_i'$.

Finally, we can compute:
\begin{align}
\frac{\partial}{\partial {u^a_k}'}{\mathbf{c}_i}' &= \frac{\partial}{\partial {u^a_k}'} 
\lbrace c_i + \sum_k \psi_k(\mathbf{c_i}) u_k \rbrace
= \psi_k(\mathbf{c_i})\, \hat{\mathbf{e}}^a
\end{align}
where $\hat{\mathbf{e}}^a$ is the unit vector along direction $a$. Projecting
this gradient onto the surface normal,
$\left\langle \frac{\partial \mathbf{c}'}{\partial {u^a_k}'}, \mathbf{n_i}'\right\rangle
= \psi_k(\mathbf{c_i})\, \hat{\mathbf{n}}_i^a$, the
full gradient evolution equation \eqref{eq:gradient_descent} yields:
\begin{align}
\frac{\partial E(u)}{\partial u^d_k} 
&= \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right]
\psi_k(\mathbf{c_i})\, \hat{\mathbf{n}}_i^a \notag\\
&2\, \underset{x,y,z}{\sum} a_d u^a_{k,d}
-2\, \underset{x,y,z}{\sum} b_d \nabla u^a_{k,d},
\label{eq:gradient_final}
\end{align}

\subsubsection{Semi-implicit Euler-forward optimization}
It is necessary to discretize $t$ in order to obtain a numerical
implementation of the equation \eqref{eq:gradient_final}:
\begin{align}
\frac{u^{t+1}-u^{t}}{\tau} 
&= - \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right]
\psi_k(\mathbf{c_i})\, \hat{\mathbf{n}}_i^a \notag\\
&-2\, \underset{x,y,z}{\sum} a_d (u^a_{k,d})^{t+1}
+2\, \underset{x,y,z}{\sum} b_d \nabla (u^a_{k,d})^{t+1}.
\end{align}

The associated Euler-Lagrange equation is found as:
\begin{equation}
(\tau^{-1} +2\, a_d - 2\, b_d \nabla )\,(u^a_{k,d})^{t+1} = 
\tau^{-1} (u^a_{k,d})^{t} - \frac{\partial E_{data}(u)}{\partial u^d_k},
\end{equation}
that is a linear system that we translate into Fourier domain,
to obtain the next deformation field $(u^a_{k,d})^{t+1}$:
\begin{multline}
(u^a_{k,d})^{t+1} = \\
 \mathcal{F}^{-1} \left\lbrace
\frac{\mathcal{F}\lbrace \tau^{-1} (u^a_{k,d})^{t} - \underset{ij}{\sum} \underset{\mathbf{c}' \in \Gamma_{ij}}{\sum}
\left[ \Delta^2_i (\mathbf{f}' ) - \Delta^2_j (\mathbf{f}')\right] \rbrace}
     {\mathcal{F}\lbrace (\tau^{-1} +2\,a_d)\mathcal{I} - 2\,b_d \Delta ) \rbrace}
     \right\rbrace
\end{multline}

@@

\subsubsection{Region descriptor reestimation}
In regular intervals, after certain number of iterations,
the parameters $\Theta_l$ of the regions can be reestimated 
based on the shifted volumetric samples 
$\mathbf{x}' = \mathbf{x_0} + u(\mathbf{x})$.

>>>>

\subsection{Proposed model}
\label{sec:methods_map}
In this paper we formulate the joint registration-segmentation
problem as follows. We look for a spatial mapping (with, typically,
$n=3$ for the 3-dimensional case):
\begin{align}
U\colon T \subset \mathbb{R}^n &\to D \subset \mathbb{R}^n \notag\\
\vec{x} &\mapsto \vec{x}' =\vec{x}+u(\vec{x}),
\label{eq:deformation_field}
\end{align}
such that the known contours in anatomical space $T$ optimally segment
the diffusion space $D$.

In a Bayesian framework, different such mappings $U$ between ``T1-''
and ``\gls*{dwi}-'' spaces are evaluated based on their posterior
probability given the observed data $F$. Using the Bayes' rule, this
posterior \emph{likelihood} can be computed as:
\begin{equation}
P(U \mid F,\Gamma_{l,m}) = \frac{P(F \mid U,\Gamma_{l,m})\, P(U)}{P(F)},
\label{eq:bayes_rule}
\end{equation}
where $P(F \mid U,\Gamma_{l,m})$ is the data-likelihood, and
$\Gamma_{l,m}$ are a set of contours in T1-space that correspond to
two interfacing regions $\Omega$, such that 
$\Gamma_{l,m} = \partial \Omega_l \cap \partial \Omega_m$.

We are interested in a mapping $\hat{U}$ which maximizes this
posterior probability (\gls*{map} estimate, \citep{bishop_pattern_2006}):
\begin{align}
\hat{U} &= \underset{U}{\argmax} \left\{ P(U \mid F,\Gamma_{l,m}) \right\} \notag\\
 &= \underset{U}{\argmax} \left\{ P(F \mid U,\Gamma_{l,m})\, P(U) \right\}.
\label{eq:map_u}
\end{align}

@@

\begin{equation}
P(F \mid U,\Gamma_{l,m}) = 
\underset{\vec{x}\in \Omega}{\prod} P\left( F(\vec{x}') \mid U(\Gamma_{l,m}) \right).
\label{eq:bayes_aposteriori}
\end{equation}

Further, within each region $\Omega_l$ defined by the contours $\Gamma$
trough duality, we assume the features to be i.i.d.:
\begin{equation}
P\left( F(\vec{x}') \mid U,\Gamma_{l,m} \right) = p_l( F(\vec{x}')) \quad
\forall \vec{x} \in \Omega_l.
\label{eq:likelihood}
\end{equation}

@@

\begin{equation}
p_l( F(\vec{x}') ) = \mathcal{N} \left( F(\vec{x}') \mid \Theta_l \right).
\label{eq:multivariate_normal}
\end{equation}

\todo{I modified this slightly}
Thus, defining the feature observed as $\vec{f}=F(\vec{x})$ and 
the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
of the descriptors of region $l$ as
\begin{equation}
\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l),
\end{equation}
we write:
\begin{align}
P( F \mid U, \Omega ) &= \underset{l}{\prod} \underset{\vec{x} \in \Omega_l}{\prod}
\mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) \notag\\
&= \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
\mdist{f'}{l} \right)}}.
\label{eq:pdf}
\end{align}

We choose a Thikonov regularization prior as follows:
\begin{equation*}
P(U) = \underset{\vec{x}}{\prod} p(u(\vec{x}))
\end{equation*}
with further:
\begin{align*}
p(u(\vec{x})) &= p_0(u(\vec{x})) \, p_1(u(\vec{x})) \notag\\
p_0(u(\vec{x})) &= \mathcal{N}( u(\vec{x}) \mid 0, A^{-1}) \notag\\
p_1(u(\vec{x})) &= \mathcal{N}( Du(\vec{x}) \mid 0, B^{-1}),
\end{align*}
that consider that the distortion and its gradient have zero
mean and variance governed by $A$ and $B$. Since the anisotropy
is technically aligned with the imaging axes, these can
be simplified:
\begin{align}
p_0(u(\vec{x})) &= \underset{\vec{x}}{\sum} \mathcal{N}( u(\vec{x}) \mid 0, (\boldsymbol{\alpha}^{\circ\frac12}\,\vec{I}_n)^{-1}) \notag\\
p_1(u(\vec{x})) &= \underset{\vec{x}}{\sum} \mathcal{N}( \nabla \cdot u(\vec{x}) \mid 0, (\boldsymbol{\beta}^{\circ\frac12}\,\vec{I}_n)^{-1})
\label{eq:priors}
\end{align}

Finally, we can turn the \gls{map} problem into 
a variational one
applying the following log-transform:
\begin{align}
E(F \mid U) &= -\log \underset{l}{\prod}
\underset{\vec{x} \in \Omega_l}{\prod} 
p_l( \vec{f}')\,p_0( u(\vec{x}))\,p_1( u(\vec{x})) \notag\\
&= C + \underset{l}{\sum}
\underset{\vec{x} \in \Omega_l}{\sum}
\mdist{f'}{l} \notag\\
&+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\alpha} \cdot u(\vec{x})^{\circ2}
+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\beta} \cdot (\nabla \cdot u(\vec{x}))^{\circ2},
\label{eq:energy}
\end{align}
that is the dual expression to the energy functional corresponding
to a discrete \gls*{acwe} framework \citep{chan_active_2001}
with anisotropic regularization as studied in
\citep{nagel_investigation_1986}.

{\color{red} {Using the region descriptors derived in \autoref{sec:methods_map}, we propose
an \gls{adf}-like, \gls{acwe}-based, piece-wise constant, image segmentation
model (where the unknown is the deformation field)
\cite{chan_active_2001} with the energy functional obtained in 
\eqref{eq:final_map_energy}. This inverse problem is ill-posed
\cite{bertero_ill-posed_1988,hadamard_sur_1902}.
In order to account for deformation field regularity and to render the 
problem well-posed, we include limiting and regularization terms into 
the energy functional \cite{morozov_linear_1975,tichonov_solution_1963}:
%
\begin{multline}
E(u) = \sum\limits_l \int_{\Omega'_l} \mdist{f}{l} \,d\vec{x} \\
+ \int \vec{u}^T \, A \, \vec{u} \, d\vec{x} + \int \tr\{(\nabla \vec{u}^T)^T B (\nabla \vec{u}^T)\} d\vec{x}
\label{eq:complete_energy}
\end{multline}
%
These regularity terms ensure that the segmenting contours in 
\gls{dwi} space are still close to their native shape. The model
easily allows to incorporate inhomogeneous and anisotropic 
regularization \cite{nagel_investigation_1986} to better regularize
the \gls{epi} distortion.}}

@@

\subsection{Numerical Implementation}
\label{sec:numerical_implementation}
%
Let us denote $\{\vec{c}_i\}_{i=1 \ldots N_c}$ the nodes of one or several shape-prior
surface(s). In our application, precise tissue interfaces of interest 
extracted from a high-resolution, anatomically correct reference volume. 
On the other hand, we have a number of \gls{dwi}-derived features at each
voxel of the volume. Let us denote by $\vec{x}$ the voxel and $F(\vec{x}) = 
[ f_1, f_2, \ldots, f_N]^T(\vec{x})$ its associated feature vector.

\subsubsection{Deformation field}
\label{sec:deformation_field}
The transformation from ``T1-'' into ``\gls{dwi}-'' coordinate space is 
achieved through a dense deformation field $u(\vec{x})$, such that:
\begin{equation}
\vec{c}_i' = U\{\vec{c}_i\} = \vec{c}_i + u(\vec{c}_i),
\label{eq:nodes_tfm}
\end{equation}
where $U$ is defined in \eqref{eq:deformation_field}. Since the nodes of the anatomical 
surfaces might lay off-grid, it is required to derive $u(\vec{x})$ from a discrete 
set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$. Densification is achieved through 
a set of associated basis functions $\psi_k$:
%
\begin{equation}
u(\vec{x}) = \sum_k \psi_k(\vec{x}) \vec{u}_k
\label{eq:intp_kernel}
\end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
of degree 3 ($B_3$).
Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
$\psi$ by the actual kernel function, the transformation writes
%
\begin{equation}
c_i' = c_i + \sum_k \left[ \vec{u}_k \, \underset{x,y,x}{\prod} B_3(c_{i,d} - \vec{x}_{k,d}) \right].
\label{eq:transformation}
\end{equation}

Based on the current estimate of the distortion $u$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$\vec{f}'$ of the regions defined by the priors in \gls{dwi} space.

\subsubsection{Gradient descent}
To find the minimum of the energy functional \eqref{eq:energy},
we propose a gradient-descent approach with respect to the underlying 
deformation field through the following partial differential equation:
\begin{equation}
\frac{\partial u(\vec{x},t)}{\partial t} \propto - \frac{\partial E(u)}{\partial u},
\label{eq:general_gradient_descent}
\end{equation}
with $t$ being an artificial time parameter of the contour
evolution, included to express the iterative character of the approach.
Now, we can introduce \eqref{eq:energy} in
\eqref{eq:general_gradient_descent}:
\begin{align}
\frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
\frac{ \partial }{\partial \vec{u}_k} \Big\{
C + \underset{l}{\sum}
\underset{\vec{x} \in \Omega_l}{\sum} \mdist{f'}{l} \notag\\
&+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\alpha} \cdot u(\vec{x})^{\circ2}
+ \underset{\vec{x} \in \Omega}{\sum} \boldsymbol{\beta} \cdot (\nabla \cdot u(\vec{x}))^{\circ2}
\Big\}.
\label{eq:gradient_descent}
\end{align}

Whereas related \glspl*{adf} introduced in \autoref{sec:methods_background}
make use of explicit level-set formulations to solve \eqref{eq:gradient_descent},
we alternatively use \emph{shape-gradients}
\cite{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}.
Let $r(\vec{x})$ be an ``arbitrary'' function over the image domain,
and $\Omega$ a certain image region with $\Gamma_{l,m}$ its corresponding
outer boundary as defined in \autoref{sec:methods_map}.
We now derive the domain integral w.r.t. $t$:
\begin{equation*}
\frac{\partial}{\partial t} \int_\Omega r(\vec{x}') d\vec{x} = \int_\Omega \frac{\partial}{\partial t}r(\vec{x}') d\vec{x} - \int_{\Gamma_{l,m}} r(\vec{x}') \left\langle \frac{\partial \Gamma_{l,m} }{\partial t}, N_{\Gamma_{l,m}}\right\rangle d\vec{x}
\end{equation*}
where $\left\langle\frac{\partial\Gamma_{l,m}}{\partial t}, N_{\Gamma_{l,m}}\right\rangle$ is 
the projection of the boundary movement on the unit inward normal $N_{\Gamma_{l,m}}$. Assuming
that the region descriptors $\Theta_l$ vary slowly enough, we can consider
that $\frac{\partial}{\partial t} r(\vec{x}') = 0$ and thus:
\begin{equation}
\frac{\partial}{\partial t} \int_\Omega r(\vec{x}') d\vec{x} = 
- \int_{\Gamma_{l,m}} r(\vec{x}') \left\langle \frac{\partial \Gamma_{l,m} }{\partial t}, N_{\Gamma_{l,m}}\right\rangle d\vec{x}
\label{eq:shape_gradients}
\end{equation}

Therefore, we can apply a discretized interpretation of \eqref{eq:shape_gradients}
to compute the data term in \eqref{eq:gradient_descent} as follows:
\begin{align}
\frac{\partial E_{data}(u)}{\partial \vec{u}_k} &=
\underset{l}{\sum} \frac{ \partial }{\partial u^d_k} \lbrace
 \underset{\vec{x} \in \Omega_l}{\sum}
{\mathcal{D}_l^2}^2 (\vec{f}') \rbrace \notag\\
&= \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right]
\left\langle \frac{\partial \vec{c}'}{\partial {u^a_k}'}, \vec{n_i}'\right\rangle,
\label{eq:gradient_wshape}
\end{align}
in this case, the formulation has been adapted to the non-binary case, $l,m$
being any pair of neighboring regions, and $\Gamma_{l,m}$ the contour separating
them such that $\vec{x}' = \vec{c}' \in\Gamma_{l,m} \iff \vec{x}\in \partial\Omega_i \cap \partial\Omega_j$
and $\vec{n_i}'$ is the unit inward normal to the contour at $c_i'$.

Finally, we can compute:
\begin{align}
\frac{\partial}{\partial {\vec{u}_k}'}{\vec{c}_i}' &= \frac{\partial}{\partial {u^a_k}'} 
\lbrace c_i + \sum_k \psi_k(\vec{c}_i) \vec{u}_k \rbrace
= \psi_k(\vec{c_i})\, \hat{\vec{e}}^a
\end{align}
where $\hat{\vec{e}}^a$ is the unit vector along direction $a$. Projecting
this gradient onto the surface normal,
$\left\langle \frac{\partial \vec{c}'}{\partial {u^a_k}'}, \vec{n_i}'\right\rangle
= \psi_k(\vec{c_i})\, \hat{\vec{n}}_i^a$, the
full gradient evolution equation \eqref{eq:gradient_descent} yields:
\begin{align}
\frac{\partial E(u)}{\partial \vec{u}_k} 
&= \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right]
\psi_k(\vec{c_i})\, \hat{\vec{n}}_i^a \notag\\
&2\, \underset{x,y,z}{\sum} a_d u^a_{k,d}
-2\, \underset{x,y,z}{\sum} b_d \nabla u^a_{k,d},
\label{eq:gradient_final}
\end{align}

\subsubsection{Semi-implicit Euler-forward optimization}
It is necessary to discretize $t$ in order to obtain a numerical
implementation of the equation \eqref{eq:gradient_final}:
\begin{align}
\frac{u^{t+1}-u^{t}}{\tau} 
&= - \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right]
\psi_k(\vec{c_i})\, \hat{\vec{n}}_i^a \notag\\
&-2\, \underset{x,y,z}{\sum} a_d (u^a_{k,d})^{t+1}
+2\, \underset{x,y,z}{\sum} b_d \nabla (u^a_{k,d})^{t+1}.
\end{align}

The associated Euler-Lagrange equation is found as:
\begin{equation}
(\tau^{-1} +2\, a_d - 2\, b_d \nabla )\,(u^a_{k,d})^{t+1} = 
\tau^{-1} (u^a_{k,d})^{t} - \frac{\partial E_{data}(u)}{\partial u^d_k},
\end{equation}
that is a linear system that we translate into Fourier domain,
to obtain the next deformation field $(u^a_{k,d})^{t+1}$:
\begin{multline}
(u^a_{k,d})^{t+1} = \\
 \mathcal{F}^{-1} \left\lbrace
\frac{\mathcal{F}\lbrace \tau^{-1} (u^a_{k,d})^{t} - \underset{l,m}{\sum} \underset{\vec{c}' \in \Gamma_{l,m}}{\sum}
\left[ \Delta^2_i (\vec{f}' ) - \Delta^2_j (\vec{f}')\right] \rbrace}
     {\mathcal{F}\lbrace (\tau^{-1} +2\,a_d)\mathcal{I} - 2\,b_d \Delta ) \rbrace}
     \right\rbrace
\end{multline}

@@

\subsubsection{Region descriptor reestimation}
In regular intervals, after certain number of iterations,
the parameters $\Theta_l$ of the regions can be reestimated 
based on the shifted volumetric samples 
$\vec{x}' = \vec{x_0} + u(\vec{x})$.
